---
output: github_document
---

See also: 

- more learning materials: https://github.com/EvaMaeRey/ay_2023_2_advanced_individual_study
- sibling project: https://github.com/EvaMaeRey/ma206equations

## Title 

Easily visualize common distributions for statistical and probabilistic analytics in ggplot2

## Abstract 

Mathematical fluency involves frequent use of probability distributions. However, visualizing these distributions can be tedious.   In a classroom setting with a given curriculum, the set of distributions to be mastered is well defined.  Having shortcuts to produce or ingest these distributions in analytic software for display and analysis could be beneficial for students and instructors alike.  The goal of this project is to provide ready-to-use dataframes and functions producing relevant dataframes.  We make these functions available in the R package `ma206distributions`.


## Which distributions are relevant for the statistics and probability course?

A good start for looking at the core of distributions in the ma206 course guide:


- binomial
- geometric
- random 

## Objectives

1. Provide additional **dataframes** for ma206 that are probabilities from curriculum (probability problems in text book).  
2. Provide *stamp* functions for binomial and geometric distributions (normal and t exist in ggxmean); also think about rewrite for stamp_normal (i.e. what should height distribution look like?)
4. Let's try:  shiny app which 1) demonstrates distribution characteristics 2) showcases new functions quoting back ggplot code
2. stretch? sampling from distributions function or guidance..



## Tidying data from discrete random variable probability problems

Tidying up data is important so that when the data is inputted into R or any coding language. This process of 'tidying' data creates a well-organized and structured format that is easier for data analysis and visualization.  

Probability tables are often communicated in untidy, wide forms not easily ingested by statistical software. (another example: http://www.stat.yale.edu/Courses/1997-98/101/ranvar.htm#:~:text=A%20discrete%20random%20variable%20is,then%20it%20must%20be%20discrete) For example, the table was provided for a probability problem about a spinning wheel with pie sectors with various prize award amounts: 

```{r, echo = F, warning=F, message=F}
library(tidyverse)
knitr::kable(t(ma206distributions::prize_wheel %>% select(-3)))
```

The table summarized the situation:

```{r, warning= F, message=F}
set.seed(1243323)
ma206distributions::prize_wheel %>% 
  uncount(frequency) %>% 
  slice_sample(n = 12) %>% 
  mutate(position = row_number()) %>% 
  ggplot() + 
  aes(x = position, y = .5, fill = payout) + 
  geom_tile(alpha = .85, color = "white") + 
  geom_text(aes(label = payout %>% paste0("$", .)), 
            color = "white", 
            y = .85,
            size = 6) + 
  coord_polar() +
  scale_fill_viridis_c(end = .8) +
  theme_void() + 
  theme(legend.position = "none") 

```

To get the data into a tidy, ready-to-use form, we transposed the table from wide to long so that a variable is a columns and the column headers inform us about the content. This way, each row forms an observation, in this case an event type. The spinning of the prize wheel has three outcomes based on which sector type the wheel landed on:  no prize, win $1, and win $3. We added a numeric column 'payout' too which we makes the data easier to use from a mathematical standpoint.  Below is our tidied data. 

```{r, echo = F}
knitr::kable(ma206distributions::prize_wheel)

```




In our package we include the data structured in this way, and call the object `prize_wheel`:

```{r}
library(ma206distributions)
print(prize_wheel)
```

We see that we can easily visualize the joint distribution of payout and frequency using the ggplot2 tool that is used heavily in the statistics part of the class.

```{r, warning=F, message=F}
library(tidyverse)
prize_wheel %>% 
  ggplot() + 
  aes(x = payout) + 
  aes(y = frequency) +
  geom_lollipop() + 
  aes(weight = frequency) +
  geom_fulcrum(color = "red", 
               shape = 2, size = 5)

```

# The tabular data can also be used to walk through computations, like finding the expected value of the prize wheel spin.

```{r}
sum(prize_wheel$payout *
  prize_wheel$frequency) /
  sum(prize_wheel$frequency)
```


```{r, warning=F, message=F}
library(tidyverse)
library(ggxmean)

ggxmean:::stamp_space() + 
  stamp_normal_dist()

```




# Project timeline

|Timing |Activity and Deliverable |Hours | 
|---|---|---|
|0-3 weeks |Background on writing functions in R and package architecture |15 | 
|3-6 weeks |ggplot2 programming basics |15 | 
|6-10 weeks |Building functionality easily visualizing probability distributions (normal, t, binomial, geometric) |15 | 
|10-12 weeks |Showcasing and soliciting feedback |15 | 
|12-15 weeks |Incorporating feedback |10 | 
|Final |Report and presentation. |10 | 



# Some initial code ideas...

Below, we do some of the visualization with base R and base ggplot2.  The aim is to provide code that feels much more effortless and fun to use to visualize distributions.

## Visualizing a binomial distribution without {ma206distributions} functions

The equations in the plot were produce in the following fashion (no functions or package pre-built): 

```{r, eval = T, warning=F, message=F}
library(tidyverse)


latex2exp::TeX("${{N}\\choose{k}} \\cdot p^kq^{N-k}$")

single_trial_prob <- .5
num_trials <- 10
0:num_trials ->
  possible_outcomes
dbinom(x = possible_outcomes,
       size = 10,
       prob = single_trial_prob) ->
  probs
tibble(possible_outcomes, probs) %>%
  ggplot() +
  aes(x = possible_outcomes) +
  scale_x_continuous(breaks = possible_outcomes) +
  aes(y = probs) +
  geom_point(fill = "white") +
  aes(xend = possible_outcomes,
      yend = 0) +
  geom_segment() +
  labs(subtitle = "Course idea #0: For multiple trials, the probability of a certain number of successes..." %>%
         str_wrap(60),
       x = "Number of Successes in 10 trials",
       y = "Probability") +
  geom_text(aes(label = probs %>% round(4)),
            nudge_y = .01) + 
    annotate(geom = "text",
    x = 1.5,
    y = .2,
    label = "{{N}*choose * {k}} %.% p^{k}*q^{N - k} ",
    parse = TRUE,
    size = 5) ->
    binomial_distribution

binomial_distribution
```


## Visualizing a Discrete Random Variable without {ma206distributions} functions

Probability problems could be structured as data frames visualized with the same tools used in the stats part of the class.

```{r}
tibble(event = c("totaled", "fender bender", "accident free"),
       probability = c(.01, .09, .9),
       payout = c(20000, 5000, 0)) ->
expected

plotmath_ev <- latex2exp::TeX("$E[X] = x_1p_1 + x_2p_2 + x_3p_3 ...x_np_n = \\sum_1^n{x_ip_i}$")

plotmath_var <- latex2exp::TeX("$Var(X) = E[(X-\\mu)^2] = \\sigma^2_x = \\sum_1^n(x_i-\\mu)^2p_i$")

plotmath_sd <- latex2exp::TeX("$\\sigma_x = SD(X) = \\sqrt{Var(X)}$")




expected %>% 
  ggplot() + 
  aes(x = payout, y = probability) + 
  geom_point() + 
  geom_segment(aes(xend = payout, yend = 0)) + 
  geom_text(aes(label = probability), size = 5, 
            vjust = -.5, color = "black") + 
  ggxmean::geom_x_mean(aes(weight = probability)) + 
  geom_vline(xintercept = sum(c(.01*20000, .09*5000)),
             linetype = "dashed", color = "black") + 
  labs(subtitle = "Course idea 8:  The expected value E.V. is the 'balancing point' of a Random variable, X.  It is the weighted average of values, (weighting values by their probabilities). In large samples from the X, the average will be close to the E.V.;" %>% str_wrap(60)) + 
  annotate(geom = "label", x =sum(c(.01*20000, .09*5000)), label = sum(c(.01*20000, .09*5000)), y = .5 ) + 
  scale_y_continuous(limits = c(0,1)) + 
  annotate(geom = "text", label = plotmath_ev, parse = T,
           x = 12000, y = .95) +
  annotate(geom = "text", label = plotmath_var, parse = T,
           x = 12000, y = .8) +  
  annotate(geom = "text", label = plotmath_sd, parse = T,
           x = 12000, y = .65) ->
discrete_random

discrete_random


```


See also:

- https://evamaerey.github.io/mytidytuesday/2022-08-01-dbinom-dgeom/dbinom_dgeom.html
- https://evamaerey.github.io/mytidytuesday/2022-08-01-dbinom-dgeom/dbinom_dgeom_flipbook.html


# Architechture we will use

### How do we bundle up our work into concept-sized usable bits? Functions.

### How do we bundle up related functions? Packages.


- We will still follow the steps outlined in this data package readme: https://github.com/EvaMaeRey/tidytitanic
- As well as consult the package building guide: https://evamaerey.github.io/package_in_20_minutes/package_in_20_minutes


### How do we collaborate and share the package? Version Control.

- create a github account for collaboration https://github.com/
- we'll make the repository available in this account

